Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-16 arxiv/proof_gaussian_case.tex
appendix

Recall that $(W-z\id)^{-1}$ is the resolvent of matrix $W$. 
We say that $(W-z\id)^{-1}$ converges to a deterministic $p\times p$ matrix limit $R(z)$ if for any sequence of deterministic vectors $v\in \R^p$ with unit norm,
$$v^\top \left[(W-z\id)^{-1}-R(z)\right]v\to 0\ \ \ \text{when $p$ goes to infinity.
}$$

@@

\paragraph{Variance asymptotic limit.}
In Theorem \ref{main_cor}, we will show that for $z$ in a small neighborhood around $0$, when $p$ goes to infinity, $G(z)$ converges to the following limit
\be \label{defn_piw}
	\Gi(z) \define \begin{pmatrix} (a_{1}(z)\Lambda^2  +  (a_{2}(z)- z)\id_{p\times p})^{-1} & 0 & 0 \\ 0 & - \frac{n_1+n_2}{n_1} a_{1}(z)\id_{n_1\times n_1} & 0 \\ 0 & 0 & -\frac{n_1+n_2}{n_2}a_{2}(z)\id_{n_2\times n_2}  \end{pmatrix},\ee
where $a_1(z)$ and $a_2(z)$ are the unique solutions to the following self-consistent equations
\be
\begin{split}\label{selfomega_a}
	&a_1(z) + a_2(z) = 1 - \frac{1}{n_1 + n_2} \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1(z) + a_2(z)}{\lambda_i^2 a_1(z) + a_2(z) - z}},  \\
	&a_1(z) + \frac{1}{n_1 + n_2}\bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1(z)}{\lambda_i^2 a_1(z) + a_2(z) - z}} = \frac{n_1}{n_1 + n_2}.
% \frac{\rho_1}{a_{1}(z)} = \frac{1}{p}\sum_{i=1}^p \frac{\lambda_i^2}{ - z+\lambda_i^2 a_{1}(z) +a_{2} (z) } + (\rho_1+\rho_2),\  \frac{\rho_2}{a_{2}(z)} = \frac{1}{p}\sum_{i=1}^p \frac{1 }{  -z+\lambda_i^2 a_{1}(z) +  a_{2}(z)  }+ (\rho_1+\rho_2) .
\end{split}
\ee
The existence and uniqueness of solutions to the above system are shown in Lemma \ref{lem_mbehaviorw}.
%First, we define the deterministic limits of $(m_1(z), m_{2}(z))$ by $\left(-\frac{\rho_1+\rho_2}{\rho_1}a_{1}(z),-\frac{\rho_1+\rho_2}{\rho_2}a_{2}(z)\right)$, where
%satisfying that $\im a_{1}(z)< 0$ and $\im a_{2}(z)<0$ for $z\in \C_+$ with $\im z$.
%\be\label{ratios}
% \gamma_n :=\frac{p}{n}=\frac{1}{\rho_1+\rho_2},\quad r_1 :=\frac{n_1}{n}=\frac{\rho_1}{\rho_1+\rho_2},\quad r_2 :=\frac{n_2}{n}=\frac{\rho_2}{\rho_1+\rho_2}.
%\ee
Given this result, we show that when $z = 0$, the matrix limit $\Gi(0)$ implies the variance limit shown in equation \eqref{lem_cov_shift_eq}.
First, we have that $a_1 = a_1(0)$ and $a_2 = a_2(0)$ since equation \eqref{selfomega_a} reduces to equation \eqref{eq_a12extra}.
Second, since $W^{-1}$ is the upper-left block matrix of $G(0)$, we have that $W^{-1}$ converges to $ (a_1\Lambda^2 + a_2\id)^{-1} $.
Using the fact that $\tr[\Sigma^{(2)} \hat{\Sigma}^{-1}] = \bigtr{W^{-1}} / (n_1 + n_2)$, we get that when $p$ goes to infinity, % the trace of $$ converges to
\begin{align*}
  \bigtr{\Sigma^{(2)} \hat{\Sigma}} \rightarrow \frac{1}{n_1+n_2}\bigtr{(a_1 \Lambda^2 + a_2\id)^{-1}} &= \frac1{n_1+n_2}\bigtr{(a_1 M^{\top}M + a_2 \id)^{-1}} \\
  &=\frac{1}{n_1+n_2} \bigtr{\Sigma^{(2)} (a_1 \Sigma^{(1)} + a_2 \Sigma^{(2)})^{-1}},
  \end{align*}
%\noindent{\bf Variance asymptotics.} Using definition \eqref{mainG}, we can write equation \eqref{eigen2extra} as
%\be\label{rewrite X as R} [(X^{(1)})^\top X^{(1)}+(X^{(2)})^\top X^{(2)}]^{-1}=n^{-1}\Sigma_2^{-1/2}V\cal G(0)V^\top\Sigma_2^{-1/2}.\ee
%When $z=0$, it is easy to check that , which means that we actually have $a_1(0)=a_1$ and $a_2(0)=a_2$. Hence the matrix limit of $\cal G(0)$ is given by $(a_{1}\Lambda^2 + a_{2}\id_p)^{-1}$. Then inserting this limit into equation \eqref{rewrite X as R}, we can write the left-hand side of equation \eqref{lem_cov_shift_eq} as
%\begin{align}
%&\bigtr{\left( (X^{(1)})^{\top}X^{(1)} + (X^{(2)})^{\top}X^{(2)}\right)^{-1} \Sigma}\approx n^{-1}\bigtr{\Sigma_2^{-1/2}V\cal (a_{1}\Lambda^2 + a_{2}\id_p)^{-1}V^\top\Sigma_2^{-1/2}\Sigma}  \nonumber\\
%&=n^{-1}\bigtr{\Sigma_2^{-1/2}\cal (a_{1}\Sigma_2^{-1/2}\Sigma_1\Sigma_2^{-1/2} + a_{2}\id_p)^{-1}\Sigma_2^{-1/2}\Sigma}  = n^{-1}\bigtr{\cal (a_{1} \Sigma_1  + a_{2}\Sigma_2)^{-1}\Sigma}  ,\label{Gi00}
%\end{align}
where we note that $M^\top M = (\Sigma^{(2)})^{-1/2} \Sigma^{(1)} (\Sigma^{(2)})^{-1/2}$ and its SVD is equal to $V^{\top}\Lambda^2 V$.
%For the asymptotic limit, its concentration error is shown in Appendix \ref{appendix RMT}.

\paragraph{Bias asymptotic limit.}
For the bias limit in equation \eqref{lem_cov_derv_eq}, we show that it is governed by the derivative of $(W - z\id)^2$ with respect to $z$ at $z = 0$.
First, we reduce the empirical bias term in equation \eqref{lem_cov_derv_eq} to $W$
\begin{align}\label{calculate G'}
	(n_1 + n_2)^2 \hat{\Sigma}^{-1}\Sigma^{(2)}\hat{\Sigma}^{-1} = {\Sigma^{(2)}}^{-1/2} V W^{-2} V^{\top} {\Sigma^{(2)}}^{-1/2}.
\end{align}
Let $\cal G(z):=(W-z\id )^{-1}$ denote the resolvent of $W$.
Our key observation is that $\frac{\dd{\cal G(z)}}{\dd z} =  \cal G^2(z)$.
Hence, provided that limit of $(W - z\id)^{-1}$ is $(a_1(z) \Lambda^2 + (a_2(z) - z) \id)^{-1}$ near $z = 0$, the limit of $\frac{\dd{\cal G(0)}}{\dd z}$ satisfies that
\begin{align}\label{cal G'0}
	\frac{\dd \cal G(0)}{\dd z} \to \frac{-\frac{\dd a_1(0)}{\dd z}\Lambda^2 - (\frac{\dd a_2(0)}{\dd z} - 1)\id}{(a_{1}(0)\Lambda^2 + a_{2}(0)\id_p)^2}.
\end{align}
To find the derivatives of $a_1(z)$ and $a_2(z)$, we take the derivatives on both sides of equation \eqref{selfomega_a}.
%\begin{align*}
%	\frac{\dd a_1(z)}{\dd z} + \frac{\dd a_2(z)}{\dd z} = -\frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{1}{\lambda_i^2 a_1 + a_2},
%	\frac{\dd a_1(z)}{\dd z} + \frac{1}{n_1 + n_2}\sum_{i=1}^p \frac{\lambda_i^2 (a_1'(z) a_2 - a_2'(z) a_1)}{(\lambda_i^2 a_1 + a_2)^2} = -\frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 a_1}{(\lambda_i^2 a_1 + a_2)^2}
%\end{align*}
Let $a_3 = - \frac{\dd a_1(0)}{\dd z}$ and $a_4 = - \frac{\dd a_2(z)}{\dd z}$.
%then taking implicit differentiation of equation \eqref{selfomega_a}
One can verify that $a_3$ and $a_4$ satisfy the self-consistent equations \eqref{eq_a34extra} (details omitted).
Applying equation \eqref{cal G'0} to equation \eqref{calculate G'}, we obtain the asymptotic limit of the bias term.
%\begin{align}
%& n^2\bignorm{\Sigma_2^{1/2} \bigbrace{ (X^{(1)})^{\top}(X^{(1)}) + (X^{(2)})^{\top}(X^{(2)}) }^{-1} \Sigma_1^{1/2} w}^2 \nonumber\\
%&\approx  w^\top \Sigma_1^{1/2}\Sigma_2^{-1/2}V\frac{a_3\Lambda^2 +(1+ a_4)\id_p}{(a_{1}\Lambda^2 + a_{2}\id)^2}V^\top \Sigma_2^{-1/2}\Sigma_1^{1/2}w= w^{\top} \Pi_\bias w, \label{calculatePibias}
%\end{align}
%where in the last step we used $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and $V \Lambda^2 V^\top=M^\top M$. This concludes equation \eqref{lem_cov_derv_eq}.

>>>>

Recall that $(W-z\id)^{-1}$ is the resolvent of matrix $W$. 
We say that $(W-z\id)^{-1}$ converges to a deterministic $p\times p$ matrix limit $R(z)$ if for any sequence of deterministic unit vectors $v\in \R^p$,
$$v^\top \left[(W-z\id)^{-1}-R(z)\right]v\to 0\ \ \ \text{when $p$ goes to infinity.
}$$

@@

\paragraph{Variance asymptotic limit.}
In Theorem \ref{main_cor}, we will show that for $z$ in a small neighborhood around $0$, when $p$ goes to infinity, $G(z)$ converges to the following limit
\be \label{defn_piw}
	\Gi(z) \define \begin{pmatrix} (a_{1}(z)\Lambda^2  +  (a_{2}(z)- z)\id_{p\times p})^{-1} & 0 & 0 \\ 0 & - \frac{n_1+n_2}{n_1} a_{1}(z)\id_{n_1\times n_1} & 0 \\ 0 & 0 & -\frac{n_1+n_2}{n_2}a_{2}(z)\id_{n_2\times n_2}  \end{pmatrix},\ee
where $a_1(z)$ and $a_2(z)$ are the unique solutions to the following self-consistent equations
\begin{align}
	&a_1(z) + a_2(z) = 1 - \frac{1}{n_1 + n_2} \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1(z) + a_2(z)}{\lambda_i^2 a_1(z) + a_2(z) - z}}, \label{selfomega_a000} \\
	&a_1(z) + \frac{1}{n_1 + n_2}\bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1(z)}{\lambda_i^2 a_1(z) + a_2(z) - z}} = \frac{n_1}{n_1 + n_2}.\label{selfomega_a}
% \frac{\rho_1}{a_{1}(z)} = \frac{1}{p}\sum_{i=1}^p \frac{\lambda_i^2}{ - z+\lambda_i^2 a_{1}(z) +a_{2} (z) } + (\rho_1+\rho_2),\  \frac{\rho_2}{a_{2}(z)} = \frac{1}{p}\sum_{i=1}^p \frac{1 }{  -z+\lambda_i^2 a_{1}(z) +  a_{2}(z)  }+ (\rho_1+\rho_2) .
\end{align}
The existence and uniqueness of solutions to the above system are shown in Lemma \ref{lem_mbehaviorw}.
%First, we define the deterministic limits of $(m_1(z), m_{2}(z))$ by $\left(-\frac{\rho_1+\rho_2}{\rho_1}a_{1}(z),-\frac{\rho_1+\rho_2}{\rho_2}a_{2}(z)\right)$, where
%satisfying that $\im a_{1}(z)< 0$ and $\im a_{2}(z)<0$ for $z\in \C_+$ with $\im z$.
%\be\label{ratios}
% \gamma_n :=\frac{p}{n}=\frac{1}{\rho_1+\rho_2},\quad r_1 :=\frac{n_1}{n}=\frac{\rho_1}{\rho_1+\rho_2},\quad r_2 :=\frac{n_2}{n}=\frac{\rho_2}{\rho_1+\rho_2}.
%\ee
Given this result, we now show that when $z = 0$, the matrix limit $\Gi(0)$ implies the variance limit shown in equation \eqref{lem_cov_shift_eq}.
First, we have that $a_1 = a_1(0)$ and $a_2 = a_2(0)$ since equations \eqref{selfomega_a000} and \eqref{selfomega_a} reduce to equations \eqref{eq_a12extra000} and \eqref{eq_a12extra} when $z=0$.
Second, since $W^{-1}$ is the upper-left block matrix of $G(0)$, we have that $W^{-1}$ converges to $ (a_1\Lambda^2 + a_2\id)^{-1} $.
Using the fact that $\tr[\Sigma^{(2)} \hat{\Sigma}^{-1}] = (n_1 + n_2)^{-1}\bigtr{W^{-1}} $, we get that when $p$ goes to infinity, % the trace of $$ converges to
\begin{align*}
  \bigtr{\Sigma^{(2)} \hat{\Sigma}} \rightarrow \frac{1}{n_1+n_2}\bigtr{(a_1 \Lambda^2 + a_2\id)^{-1}} &= \frac1{n_1+n_2}\bigtr{(a_1 M^{\top}M + a_2 \id)^{-1}} \\
  &=\frac{1}{n_1+n_2} \bigtr{\Sigma^{(2)} (a_1 \Sigma^{(1)} + a_2 \Sigma^{(2)})^{-1}},
  \end{align*}
%\noindent{\bf Variance asymptotics.} Using definition \eqref{mainG}, we can write equation \eqref{eigen2extra} as
%\be\label{rewrite X as R} [(X^{(1)})^\top X^{(1)}+(X^{(2)})^\top X^{(2)}]^{-1}=n^{-1}\Sigma_2^{-1/2}V\cal G(0)V^\top\Sigma_2^{-1/2}.\ee
%When $z=0$, it is easy to check that , which means that we actually have $a_1(0)=a_1$ and $a_2(0)=a_2$. Hence the matrix limit of $\cal G(0)$ is given by $(a_{1}\Lambda^2 + a_{2}\id_p)^{-1}$. Then inserting this limit into equation \eqref{rewrite X as R}, we can write the left-hand side of equation \eqref{lem_cov_shift_eq} as
%\begin{align}
%&\bigtr{\left( (X^{(1)})^{\top}X^{(1)} + (X^{(2)})^{\top}X^{(2)}\right)^{-1} \Sigma}\approx n^{-1}\bigtr{\Sigma_2^{-1/2}V\cal (a_{1}\Lambda^2 + a_{2}\id_p)^{-1}V^\top\Sigma_2^{-1/2}\Sigma}  \nonumber\\
%&=n^{-1}\bigtr{\Sigma_2^{-1/2}\cal (a_{1}\Sigma_2^{-1/2}\Sigma_1\Sigma_2^{-1/2} + a_{2}\id_p)^{-1}\Sigma_2^{-1/2}\Sigma}  = n^{-1}\bigtr{\cal (a_{1} \Sigma_1  + a_{2}\Sigma_2)^{-1}\Sigma}  ,\label{Gi00}
%\end{align}
where we note that $M^\top M = (\Sigma^{(2)})^{-1/2} \Sigma^{(1)} (\Sigma^{(2)})^{-1/2}$ and its SVD is equal to $V^{\top}\Lambda^2 V$.
%For the asymptotic limit, its concentration error is shown in Appendix \ref{appendix RMT}.

\paragraph{Bias asymptotic limit.}
 For the bias limit in equation \eqref{lem_cov_derv_eq}, we show that it is governed by the derivative of $(W - z\id)^2$ with respect to $z$ at $z = 0$.
First, we can express the empirical bias term in equation \eqref{lem_cov_derv_eq} as %$W$
\begin{align}\label{calculate G'}
	(n_1 + n_2)^2 \hat{\Sigma}^{-1}\Sigma^{(2)}\hat{\Sigma}^{-1} = {\Sigma^{(2)}}^{-1/2} V W^{-2} V^{\top} {\Sigma^{(2)}}^{-1/2}.
\end{align}
Let $\cal G(z):=(W-z\id )^{-1}$ denote the resolvent of $W$.
Our key observation is that $\frac{\dd{\cal G(z)}}{\dd z} =  \cal G^2(z)$.
Hence, provided that the limit of $(W - z\id)^{-1}$ is $(a_1(z) \Lambda^2 + (a_2(z) - z) \id)^{-1}$ near $z = 0$, the limit of $\frac{\dd{\cal G(0)}}{\dd z}$ satisfies that
\begin{align}\label{cal G'0}
	\frac{\dd \cal G(0)}{\dd z} \to \frac{-\frac{\dd a_1(0)}{\dd z}\Lambda^2 - (\frac{\dd a_2(0)}{\dd z} - 1)\id}{(a_{1}(0)\Lambda^2 + a_{2}(0)\id_p)^2}.
\end{align}
To find the derivatives of $a_1(z)$ and $a_2(z)$, we take the derivatives on both sides of equation \eqref{selfomega_a}.
%\begin{align*}
%	\frac{\dd a_1(z)}{\dd z} + \frac{\dd a_2(z)}{\dd z} = -\frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{1}{\lambda_i^2 a_1 + a_2},
%	\frac{\dd a_1(z)}{\dd z} + \frac{1}{n_1 + n_2}\sum_{i=1}^p \frac{\lambda_i^2 (a_1'(z) a_2 - a_2'(z) a_1)}{(\lambda_i^2 a_1 + a_2)^2} = -\frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 a_1}{(\lambda_i^2 a_1 + a_2)^2}
%\end{align*}
Let $a_3 = - \frac{\dd a_1(0)}{\dd z}$ and $a_4 = - \frac{\dd a_2(z)}{\dd z}$.
%then taking implicit differentiation of equation \eqref{selfomega_a}
One can verify that $a_3$ and $a_4$ satisfy the self-consistent equations in \eqref{eq_a34extra} (details omitted).
Applying equation \eqref{cal G'0} to equation \eqref{calculate G'}, we obtain the asymptotic limit of the bias term.
%\begin{align}
%& n^2\bignorm{\Sigma_2^{1/2} \bigbrace{ (X^{(1)})^{\top}(X^{(1)}) + (X^{(2)})^{\top}(X^{(2)}) }^{-1} \Sigma_1^{1/2} w}^2 \nonumber\\
%&\approx  w^\top \Sigma_1^{1/2}\Sigma_2^{-1/2}V\frac{a_3\Lambda^2 +(1+ a_4)\id_p}{(a_{1}\Lambda^2 + a_{2}\id)^2}V^\top \Sigma_2^{-1/2}\Sigma_1^{1/2}w= w^{\top} \Pi_\bias w, \label{calculatePibias}
%\end{align}
%where in the last step we used $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and $V \Lambda^2 V^\top=M^\top M$. This concludes equation \eqref{lem_cov_derv_eq}.
